Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-03 paper/paper.qmd
committing text changes to paper.qmd

# Introduction
The 2024 United States Presidential Election is shaping up to be a critical event, particularly as battleground states emerge as pivotal areas influencing the electoral outcome. As candidates Donald Trump and Kamala Harris intensify their campaigns, understanding voter preferences in these key regions becomes essential for predicting the election results. This paper utilizes state-level polling data to build a predictive model that estimates the likelihood of victory for each candidate in several key battleground states, including Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin.

@@

The remainder of this paper is structured as follows. In Section 2, we describe the data and variables used in the analysis. Section 3 outlines the model setup and estimation strategy. Section 4 presents the results, including the predicted election outcomes and visualizations. Finally, Section 5 discusses the implications and limitations of our findings

# Data {#sec-data}
## Overview
The dataset of the 2024 Presidential Election cycle is obtained from Project 538 (cite). The dataset is periodically updated throughout the presidential election campaign to reflect current and most up-to-date polling results. It consists of polling information from various pollsters such as the New York Times, ActiVote, Morning Consult, etc. For every poll conducted, relevant variables such as state, start_date, end_date, transparency_score, candidate_name, and percentage of support (pct) for the appropriate candidates are included. The dataset compiles data from all major pollsters in the US and provides us with an understanding of the current state of the presidential election.

The present analysis focuses on Donald Trump and Kamala Harris, the two leading candidates in the current presidential election race. For all the following variables mentioned, we collected observations for both of the candidates. We collect observations from the dataset where the ‘candidate_name’ is Trump and Harris. We are interested in the support for both candidates over time, for each state, therefore, we collect observations of ‘end_date’, and ‘state’, as well as the electorate support for each candidate is denoted as ‘pct’. Furthermore, since there are many pollsters present in the dataset, we extracted pollsters with high-quality data that is reliable, which is denoted as having a high ‘numeric_grade’ and ‘transparency_score’, details of data cleaning are further explained in [@sec-data-details]

@@

To ensure the quality of the analysis, only those polls that meet specific thresholds were included - refer to [@sec-appendix].

@@

```
For instance, @fig-summarystatistics reveals a total of 107 polls conducted, calculated as the total count of rows in our cleaned dataset. ‘Total polls’ reflects the total number of unique poll entries after filtering for only battleground states, polls with sufficient data quality, and polls that report support percentages for both Trump and Harris. The 107 polls conducted display an average support level of 47.801% for Trump and 47.999% for Harris. This suggests that based on the dataset, Trump has a leading advantage in winning the presidential election race by having a slighty higher average level of support than Harris. The highest percentage of support for Trump in the 107 polls was 51.700% while the lowest recorded support was 43%. For Harris, the highest percentage of support also reached 51.200%, with a minimum of 43.600%. These statistics illustrate the competitive landscape of the presidential election race, without a clear leading candidate in the current race.

@@

###Other Variables

@@

## Model Summary - Harris Model
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Harris Model."
#| warning: false

@@

## Model summary - Trump Model
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false

@@

## Polling Averages for Trump and Harris Over Time in Battleground States
@fit-state-polling-average provides a visual comparison of the current polling support for Donald Trump and Kamala Harris across key battleground states, tracked over time from August 2024 to October 2024. Each line plot represents the average polling percentages for both candidates in a specific state, with red lines denoting Trump’s support and blue lines representing Harris’s support. Across several states, there are noticeably increased fluctuations in polling averages for both candidates closer to the election particularly from late September to October. This pattern may indicate heightened voter interest and engagement as the election date approaches which is November 4th, 2024, it could also reflect changes in public opinion due to campaign events, political debates, or other external factors.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-polling-average
#| fig-cap: "Polling Averages for Trump and Harris Over Time in Battleground States."
#| warning: false

@@

## State-Level Probability of Trump Winning on Election Day
By incorporating our Bayesian models, we generate the predicted probabilities of winner for both Donald Trump and Kamala Harris in key battleground states, on Election Day (November 5th, 2024). @fig-state-probabilities provides a visual summary of these predicted probabilities.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

@@

## Probability of Winning by State on Election Day 
@tbl-fig-probabilities displays the numeric percentage of the probabilities of winning for Trump and Harris. By comparing the predicted probabilities, the predicted winner for each state is assigned accordingly. For instance, the table displays that Trump has a 49.48% predicted probability of winning Arizona compared to 49.24% of Harris’s, making Trump the predicted winner in Arizona. For all six other states, Harris has a higher predicted probability of winning than Trump.
```{r}
#| echo: false
#| eval: true
#| label: tbl-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

@@

## US Map Showing Predicted Winner by State on Election Day
To visualize the colour distribution of the predicted results, we emloy the US map from [@usmap] to display the results from @tbl-state-probabilities.

@@

# Print the map
print(us_map)
```
According to @fig-us-map, Arizona is the only state where Trump has a leading win while all the other six states are led by Harris. Since Harris has a six-to-one lead in the predicted percentage of support for all battleground states, we conclude that Harris is the predicted winner for the 2024 cycle of Presidential Election.

@@

## Model Performance {#sec-model-performance}

Out-of-sample testing results [Appendix @sec-model-diagnostics] demonstrate the effectiveness and reliability of our Bayesian logistic regression model in predicting state-level outcomes for the 2024 U.S. presidential election. Our accuracy test generated a 75% score for the model, indicating our model's high reliability for making predictions about state-level election outcomes. Since presidential elections hinge on state-level wins, a reliable state-by-state prediction would imply a reliable national outcome prediction.

The RMSE value generated for our model is 0.5, which indicates that the model-predicted probabilities are close to the true outcomes. The confusion matrix breaks down the model’s predictions into four categories: 1. orrectly predicted Trump wins, 2. Correctly predicted Harris wins, 3. Incorrectly predicted Trump wins, 4. Incorrectly predicted Harris wins. The generated confusion matrix for our model all had values of 0 to 2 for all four categories, indicating a balanced prediction between the two candidates, and suggesting the reliability of our model. Out-of-sample testing was conducted to test the reliability of our model, with further details provided in [Apppendix @sec-model-diagnostics].

## Weaknesses and Next Steps

While our model provides valuable insights into the electoral landscape, it faces limitations due to missing data for certain states, which results in some states being left blank in our predictions. Although our primary focus is on understanding the dynamics within battleground states, the lack of information from other states can lead to incomplete predictions which may obscure broader trends affecting voter sentiment.

@@

# Additional Data Details

>>>>

# Introduction {#sec-intro}
The 2024 United States Presidential Election is shaping up to be a critical event, particularly as battleground states emerge as pivotal areas influencing the electoral outcome. As candidates Donald Trump and Kamala Harris intensify their campaigns, understanding voter preferences in these key regions becomes essential for predicting the election results. This paper utilizes state-level polling data from Project 538 [@FiveThirtyEight] and the statistical programming language R [@citeR] to build a predictive model that estimates the likelihood of victory for each candidate in several key battleground states, including Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin.

@@

The remainder of this paper is structured as follows. In @sec-data, we describe the data and variables used in the analysis. @sec-model outlines the model setup and estimation strategy. @sec-results presents the results, including the predicted election outcomes and visualizations. Finally, @sec-discussion discusses the implications and limitations of our findings.

# Data {#sec-data}
## Overview
The dataset of the 2024 Presidential Election cycle is obtained from Project 538 [@FiveThirtyEight]. The dataset is periodically updated throughout the presidential election campaign to reflect current and most up-to-date polling results. It consists of polling information from various pollsters such as the New York Times, ActiVote, Morning Consult, etc. For every poll conducted, relevant variables such as state, start_date, end_date, transparency_score, candidate_name, and percentage of support (pct) for the appropriate candidates are included. The dataset compiles data from all major pollsters in the US and provides us with an understanding of the current state of the presidential election.

The present analysis focuses on Donald Trump and Kamala Harris, the two leading candidates in the current presidential election race. For all the following variables mentioned, we collected observations for both of the candidates. We collect observations from the dataset where the ‘candidate_name’ is Trump and Harris. We are interested in the support for both candidates over time, for each state, therefore, we collect observations of ‘end_date’, and ‘state’, as well as the electorate support for each candidate is denoted as ‘pct’. Furthermore, since there are many pollsters present in the dataset, we extracted pollsters with high-quality data that is reliable, which is denoted as having a high ‘numeric_grade’ and ‘transparency_score’, details of data cleaning are further explained in [Appendix -@sec-data-details].

@@

To ensure the quality of the analysis, only those polls that meet specific thresholds were included - for more details, refer to [Appendix -@sec-data-details].

@@

```
For instance, [@fig-summarystatistics] reveals a total of 107 polls conducted, calculated as the total count of rows in our cleaned dataset. ‘Total polls’ reflects the total number of unique poll entries after filtering for only battleground states, polls with sufficient data quality, and polls that report support percentages for both Trump and Harris. The 107 polls conducted display an average support level of 47.801% for Trump and 47.999% for Harris. This suggests that based on the dataset, Trump has a leading advantage in winning the presidential election race by having a slighty higher average level of support than Harris. The highest percentage of support for Trump in the 107 polls was 51.700% while the lowest recorded support was 43%. For Harris, the highest percentage of support also reached 51.200%, with a minimum of 43.600%. These statistics illustrate the competitive landscape of the presidential election race, without a clear leading candidate in the current race.

@@

### Other Variables

@@

## Model Summary
### Harris Model Summary
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Harris Model."
#| warning: false

@@

### Trump Model Summary
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false

@@

## Polling Averages for Trump and Harris Over Time in Battleground States
[@fit-state-polling-average] provides a visual comparison of the current polling support for Donald Trump and Kamala Harris across key battleground states, tracked over time from August 2024 to October 2024. Each line plot represents the average polling percentages for both candidates in a specific state, with red lines denoting Trump’s support and blue lines representing Harris’s support. Across several states, there are noticeably increased fluctuations in polling averages for both candidates closer to the election particularly from late September to October. This pattern may indicate heightened voter interest and engagement as the election date approaches which is November 4th, 2024, it could also reflect changes in public opinion due to campaign events, political debates, or other external factors.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-polling-average
#| fig-cap: "Polling Averages for Trump and Harris Over Time in Battleground States."
#| warning: false

@@

## State-Level Probability of Trump Winning on Election Day
By incorporating our Bayesian models, we generate the predicted probabilities of winner for both Donald Trump and Kamala Harris in key battleground states, on Election Day (November 5th, 2024). [@fig-state-probabilities] provides a visual summary of these predicted probabilities.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

@@

## Probability of Winning by State on Election Day 
[@tbl-fig-probabilities] displays the numeric percentage of the probabilities of winning for Trump and Harris. By comparing the predicted probabilities, the predicted winner for each state is assigned accordingly. For instance, the table displays that Trump has a 49.48% predicted probability of winning Arizona compared to 49.24% of Harris’s, making Trump the predicted winner in Arizona. For all six other states, Harris has a higher predicted probability of winning than Trump.
```{r}
#| echo: false
#| eval: true
#| label: tbl-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

@@

## US Map Showing Predicted Winner by State on Election Day
To visualize the colour distribution of the predicted results, we emloy the US map from [@usmap] to display the results from [@tbl-state-probabilities].

@@

# Print the map
print(us_map)
```
According to [@fig-us-map], Arizona is the only state where Trump has a leading win while all the other six states are led by Harris. Since Harris has a six-to-one lead in the predicted percentage of support for all battleground states, we conclude that Harris is the predicted winner for the 2024 cycle of Presidential Election.

@@

## Model Performance {#sec-model-performance}
Out-of-sample testing results [Appendix -@sec-model-diagnostics] demonstrate the effectiveness and reliability of our Bayesian logistic regression model in predicting state-level outcomes for the 2024 U.S. presidential election. Our accuracy test generated a 75% score for the model, indicating our model's high reliability for making predictions about state-level election outcomes. Since presidential elections hinge on state-level wins, a reliable state-by-state prediction would imply a reliable national outcome prediction.

The RMSE value generated for our model is 0.5, which indicates that the model-predicted probabilities are close to the true outcomes. The confusion matrix breaks down the model’s predictions into four categories: 1. orrectly predicted Trump wins, 2. Correctly predicted Harris wins, 3. Incorrectly predicted Trump wins, 4. Incorrectly predicted Harris wins. The generated confusion matrix for our model all had values of 0 to 2 for all four categories, indicating a balanced prediction between the two candidates, and suggesting the reliability of our model. Out-of-sample testing was conducted to test the reliability of our model, with further details provided in [Apppendix -@sec-model-diagnostics].

## Weaknesses and Next Steps
While our model provides valuable insights into the electoral landscape, it faces limitations due to missing data for certain states, which results in some states being left blank in our predictions. Although our primary focus is on understanding the dynamics within battleground states, the lack of information from other states can lead to incomplete predictions which may obscure broader trends affecting voter sentiment.

@@

# Additional Data Details {#sec-data-details}
